<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
pkgload::load_all(".", quiet = TRUE)
```

# SpaColExt

<!-- badges: start -->
[![R-CMD-check](https://github.com/Clara-Casabona/SpaColExt/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/Clara-Casabona/SpaColExt/actions/workflows/R-CMD-check.yaml)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![License: MIT](https://img.shields.io/badge/License-MIT-blue.svg)](LICENSE.md)
<!-- badges: end -->

SpaColExt is an R package for inferring extinction and colonization dates from species sighting records. It focuses on Bayesian extinction inference through time and across spatial units, with tools for:

- estimating posterior extant probabilities from sighting years;
- applying Solow-style Bayesian extinction estimators;
- evaluating posterior probabilities over a sequence of possible end years;
- applying the same workflow independently across spatial cells;
- exploring informative priors for extinction time and observation rate.

The package currently implements methods based on Solow (1993), with experimental non-homogeneous observation-rate functions inspired by Kodikara et al. (2020). Solow and Beet (2014)-style uncertain sighting support is planned but not yet part of the stable workflow.

## Installation

You can install the development version from GitHub:

``` r
# install.packages("remotes")
remotes::install_github("Clara-Casabona/SpaColExt")
```

Load the package with:

``` r
library(SpaColExt)
```

## Basic Workflow

The main input is a numeric vector of sighting years. In the simplest case, `compute_posterior_solow1993()` returns the posterior probability that the species is still extant at the end of the study interval.

```{r example-solow}
sightings <- c(1901, 1902, 1903, 1905, 1908, 1910)
start_year <- 1900
end_year <- 1920

compute_posterior_solow1993(
  sightings = sightings,
  start_year = start_year,
  end_year = end_year,
  dprior_m = solowdprior_m,
  dprior_te = solowdprior_te,
  prior = 0.5
)
```

## Posterior Probability Through Time

Use `posterior_probability_extinction_varying_end_year()` to evaluate the posterior extant probability for every year in a study period.

```{r example-varying-end-year}
posterior <- posterior_probability_extinction_varying_end_year(
  sightings = sightings,
  start_year = start_year,
  stop_year = end_year
)

data.frame(
  year = seq(start_year, end_year),
  extant_probability = posterior
)
```

## Spatial Workflow

Spatial analyses use a matrix-like object where each cell contains a vector of sighting years for one site. Cells with no observations can be set to `NA`.

```{r example-spatial}
sighting_grid <- list(
  c(1880, 1883, 1895, 1897, 1899),
  NA,
  NA,
  c(1882, 1884, 1896, 1898)
)
dim(sighting_grid) <- c(2, 2)

spatial_posterior <- spatial_posterior_probability_extinction_varying_end_year(
  sighting_grid,
  start_year = 1880,
  stop_year = 1930
)

spatial_posterior[1, 1]

plot_posterior_distribution(
  spatial_posterior,
  start_year = 1880,
  stop_year = 1930
)
```

## Informative Priors

`compute_posterior_c2022_extinction()` allows custom priors for extinction time (`prior_te`) and observation rate (`prior_m`). This is useful when independent ecological or sampling information should influence the posterior.

```{r example-informative-priors}
prior_te <- function(te) 1
prior_m <- function(m) 1 / pmax(m, .Machine$double.eps)

compute_posterior_c2022_extinction(
  sightings = sightings,
  start_year = start_year,
  stop_year = end_year,
  prior_te = prior_te,
  prior_m = prior_m
)
```

You can also compare the effect of different prior combinations:

```{r example-prior-plot}
visualize_priors_effects(
  t_start = start_year,
  t_stop = end_year,
  sightings = sightings,
  prior_te = prior_te,
  prior_m = prior_m
)
```

## Function Overview

- `compute_posterior_solow1993()`: posterior extant probability for one study interval.
- `posterior_probability_extinction_varying_end_year()`: posterior extant probabilities for a sequence of end years.
- `spatial_posterior_probability_extinction_varying_end_year()`: applies the varying end-year workflow across spatial cells.
- `plot_posterior_distribution()`: plots posterior extant probabilities through time.
- `compute_posterior_c2022_extinction()`: posterior curve with optional informative priors.
- `transform_and_reverse()`: helper for reversing sighting dates for colonization-oriented workflows.

Experimental non-homogeneous functions are available, but their interface and numerical behavior may still change.

## Development Status

SpaColExt is under active development. The current maintenance priorities are:

- add stable support for colonization-date inference;
- formalize spatial dependence between neighbouring sites;
- add support for uncertain sightings;
- expand tests around the non-homogeneous observation-rate functions;
- improve method references and vignettes for applied ecological workflows.

## References

Solow, A. R. (1993). Inferring extinction from sighting data. *Ecology*, 74(3), 962-964.

Solow, A. R. and Beet, A. R. (2014). On uncertain sightings and inference about extinction. *Conservation Biology*, 28(4), 1119-1123.

Kodikara, S., et al. (2020). Non-homogeneous sighting processes for extinction inference. *Methods in Ecology and Evolution*.
