---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# SpaColExt

<!-- badges: start -->
<!-- badges: end -->

The goal of SpaColExt R package is to infer spatial colonization and extinction dates of species accounting for the spatial heterogeneity of the data. It uses a Bayesian approach to estimate the probability of species still extant through the time a different spatial level.

We use the Bayesian functions from [Solow 1993](https://www.jstor.org/stable/1940821?seq=1) and we will also add the  [Solow & Beet 2014](https://www.researchgate.net/publication/262387815_On_Uncertain_Sightings_and_Inference_about_Extinction) functions to incorporate the certainty and uncertainty of the data. 


## Installation

You can install the development version of SpaColExt from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("Clara-Casabona/SpaColExt")
```

## Some notes:

This package is in development. I need to:

- Add spatial prior information

## Example 1 
 
This is a basic example which shows you how to estimate the posterior distribution of the extant probability of a extinct species:

```{r example}
library(SpaColExt)

## basic example code
sightings = c(1901,1902,1903,1905,1908,1910)
start_year = 1900
end_year = 1920
dprior_m = function(m) 1 / m
dprior_te = function(te) 1
prior =0.5

compute_posterior_solow1993(sightings = sightings, 
                            start_year = start_year,
                            end_year = end_year, 
                            dprior_m = dprior_m ,
                            dprior_te = dprior_te,
                            prior = prior)
```


## Example 2 
 
This is a basic example which shows you how to estimate the posterior distribution of the extant probability of a extinct species in different sites:

```{r example2}
library(SpaColExt)

## Creating data in a matrix 2x2
data = list(c(1880, 1883, 1895, 1897, 1899), NA, NA, c(1882, 1884, 1896, 1898))
dim(data) <- c(2, 2)

## Using Solow 1993 priors
dprior_m = function(m) 1 / m
dprior_te = function(te) 1
prior =0.5

## Study interval
start_year = 1880
stop_year = 1930


# Estimating extant probabilities

extant_probability = Spatial_posterior_probability_extinction_varying_end_year(data, 
                                                                               start_year=start_year,
                                                                               stop_year=stop_year)
## Output of one of the sites:

extant_probability[1,1]

# Ploting the exant probability

Plot_posterior_distribution(extant_probability, start_year=start_year, stop_year=stop_year)
```


With this example we estimate the probability that the species is extant in different sites considering that the observations are independent between the sites.

## Example 3  (In development)

This  example will show you how to estimate the posterior distribution of the extant probability of an extinct species in different sites, assuming that the observations in one site are influenced by the  observations in neighbours sites:

```{r example3}
library(SpaColExt)

## Creating data in a matrix 4x4
data = list(c(1880, 1883, 1895, 1897, 1899), NA, NA, c(1882, 1884, 1896, 1898),NA,c(1880, 1883, 1895, 1897, 1899),c(1880, 1883, 1895, 1897, 1899),c(1880, 1883, 1895, 1897, 1899),c(1880, 1883, 1895, 1897, 1899),NA, NA, NA, c(1880, 1883, 1895, 1897, 1899), NA, NA, NA )
dim(data) <- c(4, 4)

## Using spatial priors
#dprior_m = # LGCP with inla 
#dprior_te = function(te) 1
#prior =0.5

## Study interval
#start_year = 1880
#stop_year = 1930


# Estimating extant probabilities

#extant_probability = Spatial_posterior_probability_extinction_varying_end_year(data, 
#                                                                               start_year=start_year,
#                                                                               stop_year=stop_year)
## Output of one of the sites:

#extant_probability[1,1]

# Ploting the exant probability

#Plot_posterior_distribution(extant_probability, start_year=start_year, stop_year=stop_year)

# Spatial map by year

```



