# Spatial, Colonization, and Priors

## Spatial Workflow

Spatial analyses use a matrix-like object where each cell contains a vector of sighting years for one site. Cells with no observations can be set to `NA`.

By default, spatial cells are treated independently. SpaColExt also includes a simple neighbour-informed adjustment for cases where adjacent cells are expected to have similar extinction or persistence dynamics. This is not a full Bayesian spatial hierarchical model. It is a transparent spatial smoothing step applied after local posterior curves have been estimated:

$$
\text{logit}(p_i^{spatial}(t)) =
(1 - \rho)\text{logit}(p_i(t)) +
\rho \frac{1}{|N_i|}\sum_{j \in N_i}\text{logit}(p_j(t))
$$

where \(p_i(t)\) is the local posterior probability in cell \(i\), \(N_i\) is the set of neighbouring cells, and \(\rho\) controls neighbour influence.

```{r}
pkgload::load_all("..", quiet = TRUE)

sighting_grid <- list(
  c(1880, 1883, 1895, 1897, 1899), c(1881, 1884, 1892, 1901), NA, NA,
  c(1882, 1888, 1894, 1898), c(1885, 1892, 1902, 1910), c(1887, 1899, 1905), NA,
  NA, c(1890, 1895, 1901), c(1892, 1898, 1904, 1912), c(1894, 1902, 1914),
  NA, c(1898, 1904), c(1901, 1908, 1916), c(1902, 1910, 1920)
)
dim(sighting_grid) <- c(4, 4)

spatial_posterior <- spatial_posterior_probability_extinction_varying_end_year(
  sighting_grid,
  start_year = 1880,
  stop_year = 1930
)

spatial_posterior_smoothed <- smooth_spatial_posterior(
  spatial_posterior,
  rho = 0.35
)
```

```{r}
independent_grid <- final_year_probability_grid(spatial_posterior)
smoothed_grid <- final_year_probability_grid(spatial_posterior_smoothed)

grid_df <- rbind(
  data.frame(
    row = rep(seq_len(nrow(independent_grid)), times = ncol(independent_grid)),
    col = rep(seq_len(ncol(independent_grid)), each = nrow(independent_grid)),
    extant_probability = as.vector(independent_grid),
    model = "Independent cells"
  ),
  data.frame(
    row = rep(seq_len(nrow(smoothed_grid)), times = ncol(smoothed_grid)),
    col = rep(seq_len(ncol(smoothed_grid)), each = nrow(smoothed_grid)),
    extant_probability = as.vector(smoothed_grid),
    model = "Neighbour-informed"
  )
)

ggplot2::ggplot(grid_df, ggplot2::aes(col, row, fill = extant_probability)) +
  ggplot2::geom_tile(color = "white", linewidth = 0.8) +
  ggplot2::geom_text(
    ggplot2::aes(label = ifelse(is.na(extant_probability), "", sprintf("%.2f", extant_probability))),
    size = 3
  ) +
  ggplot2::scale_y_reverse(breaks = seq_len(4)) +
  ggplot2::scale_x_continuous(breaks = seq_len(4)) +
  ggplot2::scale_fill_viridis_c(
    option = "C",
    limits = c(0, 1),
    na.value = "grey92",
    name = "P(extant)"
  ) +
  ggplot2::coord_equal() +
  ggplot2::facet_wrap(~model) +
  ggplot2::theme_bw(base_size = 10) +
  ggplot2::theme(panel.grid = ggplot2::element_blank()) +
  ggplot2::labs(x = "Column", y = "Row")
```

These grids summarize the posterior probability of persistence in the final year of the study period. The neighbour-informed version partially pools each cell with its rook neighbours. This can stabilize isolated cells with sparse observations, but it should be interpreted as spatial smoothing rather than proof of dispersal or local demographic coupling.

## Colonization by Time Reversal

Colonization can be treated as the temporal mirror of extinction. If a species is absent before colonization and present after colonization, reversing the time axis turns the problem into an extinction-like problem. SpaColExt uses this idea in `posterior_probability_colonization_varying_year()`.

```{r}
set.seed(42)

colonization_start_year <- 1980
colonization_stop_year <- 2020
true_colonization_year <- 1998

colonization_sightings <- sort(unique(
  true_colonization_year + floor(cumsum(stats::rexp(20, rate = 0.35)))
))
colonization_sightings <- colonization_sightings[colonization_sightings <= colonization_stop_year]

colonization_posterior <- posterior_probability_colonization_varying_year(
  sightings = colonization_sightings,
  start_year = colonization_start_year,
  stop_year = colonization_stop_year
)

colonization_df <- data.frame(
  year = seq(colonization_start_year, colonization_stop_year),
  colonization_probability = colonization_posterior
)

ggplot2::ggplot(colonization_df, ggplot2::aes(year, colonization_probability)) +
  ggplot2::geom_line(linewidth = 0.8, color = "#009E73") +
  ggplot2::geom_point(size = 1.6, color = "#009E73") +
  ggplot2::geom_vline(xintercept = true_colonization_year, linetype = "dashed", color = "grey40") +
  ggplot2::coord_cartesian(ylim = c(0, 1)) +
  ggplot2::theme_bw(base_size = 10) +
  ggplot2::labs(x = "Year", y = "Posterior colonization probability")
```

## Informative Priors

`compute_posterior_c2022_extinction()` allows custom priors for extinction time (`prior_te`) and observation rate (`prior_m`). This is useful when independent ecological or sampling information should influence the posterior.

```{r}
sightings <- c(1901, 1902, 1903, 1905, 1908, 1910)
start_year <- 1900
end_year <- 1920

prior_te <- function(te) stats::dnorm(te, mean = 1913, sd = 3)
prior_m <- function(m) stats::dlnorm(
  pmax(m, .Machine$double.eps),
  meanlog = log(2),
  sdlog = 0.45
)

compute_posterior_c2022_extinction(
  sightings = sightings,
  start_year = start_year,
  stop_year = end_year,
  prior_te = prior_te,
  prior_m = prior_m
)
```

```{r}
visualize_priors_effects(
  t_start = start_year,
  t_stop = end_year,
  sightings = sightings,
  prior_te = prior_te,
  prior_m = prior_m
)
```
